Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-04-08 paper.tex
pop studies improved

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.99\columnwidth]{figures/Fermi.comp.CR.diff.eps}
  \caption{\it Fermi-LAT flux upper limits are contrasted to the predictions by
    the semi-analytic Pinzke and Pfrommer model for CRp induced gamma-ray
    emission. We show the mean differential flux in the energy range
    $E_\gamma=1-10$~GeV for 32 clusters. The Fermi upper limits are shown with
    black arrows. The blue boxes show the gamma-ray emission from CRp induced
    $\pi^0$-decay, where the upper bound shows the estimates for an optimistic
    model and the lower bound a more conservative model excluding point sources
    (note that both models assume the maximally allowed acceleration efficiency
    for strong shocks and most likely will have to be corrected down, see
    \cite{2010MNRAS.409..449P} for details). Note that there is currently no
    tension between our predictions and upper limits from Fermi, although Virgo,
    Norma, and Coma are coming close and will enforce constraints on the
    hadronic models in the coming years.}
 \label{fig15}
\end{minipage}
\end{figure*}

the tightest limits on DM and CRps from Fermi are in the 1-10 GeV
 regime, where Fornax, M49, and NGC4636 are the most constraining
 "clusters" for the DM and Virgo, Coma and Norma are very close to the
 predictions for the CRps

using beta profiles, discrepancy with Norma cluster in Ackermann gone,
Integrated flux similar

AGN M87, effective area better in 1 GeV range

\begin{figure*}%[t]
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.99\columnwidth]{figures/Flux.comp.CR.eps}
 \caption{\it Comparison of the gamma-ray flux induced by CRps for the 106
   clusters included in the extended HIFLUGCS catalogue. The fluxes are
   calculated within $\rvir$, and are derived using a single beta profile for
   each cluster's gas density profile as obtained by ROSAT X-ray
   observations. The CR distribution follows the semi-analytically model derived
   by hydrodynamical cosmological cluster simulations \cite{2010MNRAS.409..449P}
   (see text for details). The upper panel shows the energy integrated flux
   above 100~MeV (left side) and above 100~GeV (right side), both as a function
   of HIFLUGCS cluster ID. The eight brightest clusters are labelled. The lower
   panel shows the relative difference between the gamma-ray flux predicted by
   the mass-luminosity scaling relation and to the flux computed using the
   semi-analytical model \cite{2010MNRAS.409..449P}. The name of the clusters
   with the largest offset are printed out, which correlate with cool core
   clusters.}
 \label{fig19}
\end{minipage}
\end{figure*}
 best clusters, stacking unlikely to improve

>>>>

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.99\columnwidth]{figures/Fermi.comp.CR.diff.eps}
  \caption{\it Fermi-LAT flux upper limits are contrasted to the predictions by
    the semi-analytic Pinzke and Pfrommer model for CRp induced gamma-ray
    emission. We show the mean differential flux in the energy range
    $E_\gamma=1-10$~GeV for 32 clusters. The Fermi upper limits are shown with
    black arrows. The blue boxes show the gamma-ray emission from CRp-induced
    $\pi^0$-decay, where the upper bound shows the estimates for an optimistic
    model and the lower bound a more conservative model excluding point sources
    (see \cite{2010MNRAS.409..449P} for details). Note that our models are in
    perfect agreement with the derived upper limits from Fermi. Interestingly,
    the limits of Virgo, Norma, and Coma are closing in on our predictions and
    will enforce constraints on the hadronic models in the coming years.}
 \label{fig15}
\end{minipage}
\end{figure*}

In Fig.~\ref{fig15}, we compare Fermi-LAT flux upper limits to the flux
predictions by the semi-analytical Pinzke and Pfrommer model
\cite{2010MNRAS.409..449P} for CRp induced $\gamma$-rays that uses the
X-ray-inferred density profiles of clusters as an input. For consistency, we use
beta-model fits to ROSAT data of X-ray bright clusters for our $\gamma$-ray flux
estimates \cite{2002ApJ...567..71}). Hence these predictions are superior to
those that use luminosity-mass scaling relation to infer the expected
$\gamma$-ray flux---a method used by reference \cite{2010ApJ...717L..71A} to
compare to Fermi limits.  In this semi-analytic model, we employ the fact that
CRs accelerated at structure formation shocks and transported over cosmic time
exhibit a universal spectral form and an almost universal spatial
distribution. Depending on whether or not we account for the bias of
``artificial galaxies'' in cosmological simulations, we derive an optimistic or
conservative limit of the expected $\gamma$-ray emission from decaying pions
(bracketing our ignorance of the contribution of cluster galaxies to the overall
$\gamma$-ray flux from a cluster). Note that both models assume the maximally
allowed acceleration efficiency for strong shocks that transfer 50\% of the
shock kinetic energy to CRs and most likely will have to be corrected down if
this number is not realized at structure formation shocks. Secondly, these
simulations neglect active CR transport such as streaming and diffusion relative
to the gas, i.e. we assume that advective transport is the dominant one and CRs
are tightly coupled to the gas via flux-frozen magnetic fields tangled on
sufficiently small scales. Evidence from the statistics of radio halos suggests
that this might not be the case for relaxed clusters, arguing that CR streaming
in those objects could be responsible for a decrease in the central CR number
density. This effect would also reduce the expected $\gamma$-ray flux by up to a
factor of five \cite{2011A&A...527A..99E}. We note that the tightest Fermi
limits are obtained in the 1-10 GeV regime due to the largest effective area of
LAT there. In this energy band, the Fermi limits come close to the flux
predictions of Virgo, Coma and Norma. Using integrated flux limits from Fermi,
we arrive at very similar results.  We emphasize that by using our semi-analytic
modeling, the slight discrepancy of the scaling relation-based prediction with
the Fermi limit on Norma \cite{2010ApJ...717L..71A} is resolved: Norma simply
seems to be less bright than an average cluster of the same mass as Norma. The
insenstitive limit on Perseus comes about since the central AGN in NGC1275 has
turned on over the last years, making it hard to determine a comparable
limit as obtained from teh other clusters.

\begin{figure*}%[t]
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.99\columnwidth]{figures/Flux.comp.CR.eps}
 \caption{\it Comparison of the gamma-ray flux induced by CRps for the 106
   clusters included in the extended HIFLUGCS catalogue. The fluxes are
   calculated within $\rvir$, and are derived using a single beta profile for
   each cluster's gas density profile as obtained by ROSAT X-ray
   observations. The CR distribution follows the semi-analytically model derived
   by hydrodynamical cosmological cluster simulations \cite{2010MNRAS.409..449P}
   (see text for details). The upper panel shows the energy integrated flux
   above 100~MeV (left side) and above 100~GeV (right side), both as a function
   of HIFLUGCS cluster ID. The eight brightest clusters are labelled. The lower
   panel shows the relative difference between the gamma-ray flux predicted by
   the mass-luminosity scaling relation and to the flux computed using the
   semi-analytical model \cite{2010MNRAS.409..449P}. The name of the clusters
   with the largest offset are printed out, which correlate with cool core
   clusters.}
 \label{fig19}
\end{minipage}
\end{figure*}
